---
title: "How Flights Behaved in July 2018: Delays, Cancellations, and Patterns"
author: "Kye Toussaint"
date: "2025-12-01"
format:
  html:
    theme: cosmo
    toc: true
    code-fold: true
    number-sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 4
)

library(tidyverse)
library(lubridate)
library(scales)
library(forcats)
```

```{r load-data}

q1 <- readr::read_csv("data/Q1.csv", show_col_types = FALSE)
q2 <- readr::read_csv("data/Q2.csv", show_col_types = FALSE)
q3 <- readr::read_csv("data/Q3.csv", show_col_types = FALSE)
q4 <- readr::read_csv("data/Q4csv.csv", show_col_types = FALSE)
q5 <- readr::read_csv("data/Q5csv.csv", show_col_types = FALSE)
q6 <- readr::read_csv("data/Q6csv.csv", show_col_types = FALSE)
q6b <- readr::read_csv("data/Q6bcsv.csv", show_col_types = FALSE)
q7 <- readr::read_csv("data/Q7csv.csv", show_col_types = FALSE)
```

## 1. Big Picture: What Story Are We Telling?

In this project, we will be using SQL to summarize a month of U.S. flight activity and then build visualizations using R to answer:

- **Which airlines suffer the worst delays?**
- **How do flight volumes vary across days of the week?**
- **Which airports are delay and cancellation “hot spots”?**
- **How do daily flight counts evolve over time?**

**Air travel demand is highly uneven across the week, a small number of carriers and airports experience extreme delays, and cancellations are concentrated at major hubs where disruptions ripple through the entire network.**

---

## 2. Loading Data 

Our SQL outputs already summarize the raw flights table, but some of the CSVs may include header rows or artifacts. We’ll clean those up here.

```{r clean-data}
# Q1: Max departure delay by airline
q1_clean <- q1 |>
  dplyr::filter(airline != "Reporting_Airline")

# Q2: Max early departure minutes by airline
q2_clean <- q2 |>
  dplyr::filter(airline != "Reporting_Airline")

# Q3: Flights by day of week (drop any NA / artifact row)
q3_clean <- q3 |>
  dplyr::filter(!is.na(day_name)) |>
  dplyr::mutate(
    # Use the rank from SQL to order bars from busiest to least busy
    day_name = forcats::fct_reorder(day_name, day_rank)
  )

# Q4: Worst airport by average departure delay (already filtered in SQL)
q4 <- q4

# Q5: Airline + airport combinations with high average departure delays
q5_clean <- q5

# Q6: Total number of cancelled flights in the period
total_cancelled <- q6$num_cancelled_flights[1]

# Q6b: per-airport cancelled flights with reason code
#   A = Carrier, B = Weather, C = National Airspace System (NAS)
q6b_clean <- q6b

# Q7: Daily flight counts and 3-day moving average
q7_clean <- q7 |>
  dplyr::filter(FlightDate != "FlightDate") |>
  dplyr::mutate(
    FlightDate = lubridate::ymd(FlightDate)
  )
```


## 3. Airline Extremes: Who Has the Worst Delays?

### 3.1 Maximum departure delay by airline

```{r airline-max-delay}
q1_clean |>
  ggplot(aes(x = reorder(airline, max_dep_delay),
             y = max_dep_delay)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Worst Departure Delays by Airline",
    subtitle = "Maximum departure delay (in minutes) observed in the data",
    x = "Airline (carrier code)",
    y = "Max departure delay (minutes)"
  ) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal(base_size = 12)
```

```{r airline-delay-summary, echo=FALSE}
q1_clean |>
  dplyr::arrange(dplyr::desc(max_dep_delay)) |>
  dplyr::slice_head(n = 5)
```

**Narrative:**

- Some carriers have extreme worst-case delays of many hours.
- Even if these are rare events, they highlight how **a single severely delayed flight can cascade** into missed connections and schedule disruptions.

---

### 3.2 Early departures: who leaves too soon?

```{r airline-max-early}
q2_clean |>
  ggplot(aes(x = reorder(airline, max_early_departure_minutes),
             y = max_early_departure_minutes)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Maximum Early Departures by Airline",
    subtitle = "How early the earliest flight left, by airline",
    x = "Airline (carrier code)",
    y = "Max early departure (minutes)"
  ) +
  theme_minimal(base_size = 12)
```

```{r airline-early-summary, echo=FALSE}
q2_clean |>
  dplyr::arrange(dplyr::desc(max_early_departure_minutes)) |>
  dplyr::slice_head(n = 5)
```

**Narrative:**

- The same carriers that show large delays can also leave substantially early.
- Early departures still matter: passengers arriving “on time” can miss flights if they board early.

---

##4. When Do We Fly? Day-of-Week Flight Patterns

```{r flights-by-day}
q3_clean |>
  ggplot(aes(x = day_name, y = num_flights)) +
  geom_col() +
  labs(
    title = "Total Flights by Day of Week",
    subtitle = "Busiest to least busy days based on SQL flight counts",
    x = "Day of week",
    y = "Number of flights"
  ) +
  scale_y_continuous(labels = scales::comma) +
  theme_minimal(base_size = 12)
```

```{r flights-by-day-table, echo=FALSE}
q3_clean
```

**Key insight:**

- Some weekdays carry substantially more flights than weekends, which has implications for staffing and capacity planning at airports.

---

## 5. what was the Worst Delays?

### 5.1 Single worst airport by average delay

```{r worst-airport}
q4
```

**Narrative:**

- This table highlights the airport with the **highest average departure delay**.
- On average, flights there are significantly late, signaling reliability issues.

### 5.2 Airline–airport combinations with long averages

```{r worst-airline-airport}
q5_top10 <- q5_clean |>
  dplyr::arrange(dplyr::desc(avg_dep_delay)) |>
  dplyr::slice_head(n = 10)

q5_top10
```

```{r plot-worst-airline-airport}
q5_top10 |>
  dplyr::mutate(
    label = paste(airline, "at", airport_name),
    label = forcats::fct_reorder(label, avg_dep_delay)
  ) |>
  ggplot(aes(x = label, y = avg_dep_delay)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top 10 Airline–Airport Combinations by Average Departure Delay",
    x = "Airline at Airport",
    y = "Average departure delay (minutes)"
  ) +
  theme_minimal(base_size = 11)
```

**Narrative:**

- Some combinations have average delays around or above an hour, suggesting **systematic route-level issues**, not just random noise.

---

## 6. Cancellations

### 6.1 Overall cancellations

```{r total-cancellations}
total_cancelled
```

There are **`r total_cancelled` cancelled flights** in this period.

### 6.2 top airport Cancellations

```{r cancellations-by-airport}
canc_by_airport <- q6b_clean |>
  dplyr::group_by(airport_name) |>
  dplyr::summarise(
    total_cancelled = sum(num_cancelled),
    .groups = "drop"
  ) |>
  dplyr::arrange(dplyr::desc(total_cancelled)) |>
  dplyr::slice_head(n = 10)

canc_by_airport
```

```{r plot-cancellations-by-airport}
canc_by_airport |>
  dplyr::mutate(airport_name = forcats::fct_reorder(airport_name, total_cancelled)) |>
  ggplot(aes(x = airport_name, y = total_cancelled)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top 10 Airports by Number of Cancellations (Subset)",
    x = "Airport (city, state)",
    y = "Cancelled flights"
  ) +
  theme_minimal(base_size = 11)
```




## 7. Why This Story Matters

- **Delays and cancellations are not random; they are concentrated in specific carriers, routes, and airports.**
- **Demand is uneven across the week and across days, which should guide staffing and capacity decisions.**
- By combining **SQL summaries** and **visual storytelling** in Quarto, we turn raw flight data into actionable insight for operations, policy, and passenger experience.
